Historical trends in the contribution of six major causes of death to the black-white

### 2017-04-25 Manuscript/Manuscript_COD_Trends.Rmd

The United States has shown improvement in life expectancy for both blacks and whites since 1969. For males, much of this improvement was due to narrowing differences in injury-related mortality, but these contributions were rivalled by an increasing gap in CVD-related mortality, and point to the cause-specific interventions necessary to decrease the black-white gap further. In women, a crossover in injury-related mortality led to a narrower gap that was realized partially by increasing injury-related mortality among whites. This work emphasizes the importance of understanding how the direction of changes in mortality by race results in changes to the life expectancy gap.

@@

The life expectancy of a population is a barometer for overall health. In 2014, life expectancy in the United States was 78.8 years, but hidden within this summary are important differences by subgroups, such as self-defined racial ethinicity. Whites lived 3.4 years longer than Blacks [ref: http://www.cdc.gov/nchs/data/nvsr/nvsr65/nvsr65_04.pdf]. While both black and white Americans have generally experienced increases in life expectancy since the beginning of the 20^th^ century, the difference between their life expectancies has varied.

@@

These models were run separately for every combination of state, race, and sex. They were fit using two markov chains of length 10,000. One thousand iterations remained after discarding the first two thousand iterations as burn-in and sampling every eighth iteration (to reduce autocorrelation). Each iteration produced death count estimates that were smoothed over time for every cause of death within each age group, which were used to estimate life expectancy, the life expectancy gap, and perform a cause of death decomposition. These 1,000 decompositions form the posterior distribution from which the credible intervals were estimated using the 2.5^th^ and 97.5^th^ percentiles.

@@

We used random effects meta regression to describe the general pattern in these trend lines. Each state's mean contribution was regressed against a linear spline for calendar time that was estimated using four internal knots. The model included random effects for state and was inversely-weighted by the standard error of each state's mean estimate within each year. To identify divergent states, we compared the state-specific trends (estimated separately for each state using the Bayesian model) to the general pattern estimated using meta regression.

@@

Cardiovascular disease (CVD) is the most common cause of death for both men and women of both races. Mortality from CVD decreased markedly since 1969 for both sexes and races (**eFigures 1 and 2**). However, white males underwent relatively larger mortality decreases than black males, resulting in a modest increase to the contribution of CVD to the life expectancy gap for most states (**Figure 3**). For females, black women experienced relatively larger mortality decreases over time compared to white females, resulting in a decrease to the CVD contribution to the life expectancy gap that was greater than one year in several states (**Figure 4**). For both genders, Michigan, Wisconsin, California, and DC exhibit recent contributions that are relatively larger than the general pattern, while Massachusettes and New York exhibit persistently lower levels of CVD contributions.

@@

For both races and both sexes, mortality from cancer peaked in the mid 1990s and decreased thereafter (**eFigures 3 and 4**). Among males, the peak mortality rate was much higher among blacks, which resulted in an inverted-U shape pattern in the difference in mortality between black and white males and in cancer's contribution to the life expectancy gap (**Figure 5**). For females, while cancer-related mortality continues to be higher among blacks in many states, the inverted-U pattern in mortality is similar between blacks and whites resulting in a flat general pattern in the contribution of cancer to the life expectancy over time (**Figure 6**). For both genders, DC and Wisconsin exhibit the largest cancer contributions (and also exhibit high initial values among males, but not females), while New York and Massachusettes underwent large decreases in the amount that cancer contributed to the life expectancy gap.

@@

Mortality from non-communicable disease has been increasing over time for black and white males in most states, with DC as a notable exception (**eFigure X**). Recently, there has been a slight decrease in mortality rates, especially among black males. These trends are similar among females, except that black females exhibited a sharper decline in mortality in recent years, eliminating the mortality difference in several states (**eFigure Y**). As such, the contribution of non-communicable disease to the mortality gap was generally flat among men with a recent decrease, and exhibited a gradual decrease over time among women (**Figures 7 and 8**). As with previous causes, New York and Massachusettes exhibit decreasing contributions, and Nevada has also made substantial progress. Furthermore, while DC's mortality rates decreased, they still have the highest contribution of non-communicable disease to the gap.

@@

Communicable disease is the least most common cause of death, although the 1990s exhibited an enormous increase in mortality related to communiable disease at the height of the HIV/AIDS epidemic before anti-retroviral therapy was available. This spike was greater among males, especially blacks, and within states with larger urban centers. Thus, the contribution of communicable disease to the life expectancy gap exhibits a peak in the mid 1990s, with states diverging from this general pattern according to the intensity of their local epidemics (**Figures 11 and 12**). DC still exhibits higher communicable disease-related contributions among men, while New Jersey and Florida additionally show higher contributions among females.

@@

Mortality from all other causes generally decreased earlier in the time period, especially among blacks. White men and women have shown increasing mortality rates from all other causes since the 1980s, along with black women, while black men have continued to experience decreasing rates within several states. As such, the contribution of all other causes to the life expectancy gap has generally decreased over time, and reached equality in New York, Massachusettes, and Rhode Island for both genders (**Figures 13 and 14**). The most disparate states include Pennsylvania and DC, with much larger contributions estimated for all other causes in the earlier years within Pennsylvania and in the later years within DC.

@@

###Discussion

@@

- While we found that many state-level trends followed the same pattern of COD-specific contributions, several states diverged in ways that either resulted in the near elimination of the life expectancy gap, or a less favorable result. New York and Massachusettes have virtually eliminated the black-white life expectancy gap for both men and women. While these states showed contribution decreases for all major causes of death except communicable disease, both states began with only a minimal contribution of CVD to the mortality gap.

@@

We used data over 45 years, across two race/ethnicities, and for all age groups. By examining a wide time window, trends were examined that provided context for the contributions estimated in the most recent year. This allowed us to recognize when the estimated contributions for a particular cause were relatively low *because they started low* vs. low *because they had decreased over time*. Furthermore, our analysis was centred on decomposing life expectancy at birth, a measure that integrates mortality rates across all ages. This is especially important in light of recent studies that have recieved criticism for narrowly focusing on a specific age group, which may present a distorted reality [Case & Deaton, Minton IJE 2016, ]. Alongside the decomposition results and age-standardized mortality rates, the scope of our analysis allows readers to examine differences in state-level inequality from many angles. [Methodologically, using the Bayesian autoregressive model, allowed us to estimate mortality rates in strata that were sparsely populated...*something about methods?*]

Our analysis has two main limitations. Firstly, we did not account for hispanic ethinicity, implying that both black and white populations in our study include some portion of hispanics that is known to differ over time and by state. We decided against excluding or adjusting for hispanic ethnicity in this study because the indicator for ethnicity has been noted to have varying accuracy across states and during the earlier time period [ref]. Thus, states with demographic shifts due to immigration from Mexico such as Florida, California, and Texas, will have estimates of life expectancy that are likely higher than they would have been had hispanics been excluded [ref]. In a previous sensitivity analysis, we found that accounting for hispanic ethinicity did not have a substantial effect on the results [ref Harper 2012], providing some evidence that not accounting for hispanic ethnicity should not substantially affect our findings.

Secondly, our analysis does not account for immigration, or migration between states. Individuals who immigrate to the US have been shown to have life expectancies at birth that are 5.8 and 4.8 years longer for men and women, respectively in 2008-2010 [ref Preston 2014]. As we are examine the difference in life expectancy between blacks and whites, this gap will be affected by the amount of immigration to the extent that new immigrants identify as black or white. Migration between states can also affect life expectancy. Migration is particularly important for the District of Columbia, because of it's small geographical bounds (it is easier to be considered migration when regions are smaller - how to say this?), and substantial in-migration from New York, in- and out-migration to contiguous Maryland and Virginia. Middle and upper-class black residents have been shown to move to the suburbs in Maryland and Virginia, while affluent whites are imigrating to DC. These changes would both serve to increase the racial gap in DC.

helpful: https://www.nytimes.com/2014/08/16/upshot/mapping-migration-in-the-united-states-since-1900.html
-movement of southern blacks to northern states, and migration out of rust belt.
-DC - influx in immigrants since 1980s, huge migration from other US states (NY, MD then Virginia), but also huge outflux to MD and Virginia. Should almost consider these three places as one.
-lots of migraiton into MD
-NY - 24% are immigrants
-Texas has influx of immigrants to 17%
-lots of migration into virginia, as well as immigration
-Florida is the state most known for in-migration, both from states (Northeast particularly NY) and from outside the US
-Alabama gets migrants from other southern states
-Connecticut - influx in those immigrating from outside US since 1990s

@@

Read: racial disparitiy in where people from DC move:
-https://ggwash.org/view/38326/theres-a-big-racial-disparity-in-where-people-born-in-dc-end-up-moving-or-not-moving

>>>>

The United States has shown improvement in life expectancy for both blacks and whites since 1969. For males, much of this improvement was due to narrowing differences in injury-related mortality, but these contributions were rivaled by an increasing gap in CVD-related mortality, and point to the cause-specific interventions necessary to decrease the black-white gap further. In women, a crossover in injury-related mortality led to a narrower gap that was realized partially by increasing injury-related mortality among whites. This work emphasizes the importance of understanding how the direction of changes in mortality by race results in changes to the life expectancy gap.

@@

The life expectancy of a population is a barometer for overall health. In 2014, life expectancy in the United States was 78.8 years, but hidden within this summary are important differences by subgroups, such as self-defined racial ethnicity. Whites lived 3.4 years longer than Blacks [ref: http://www.cdc.gov/nchs/data/nvsr/nvsr65/nvsr65_04.pdf]. While both black and white Americans have generally experienced increases in life expectancy since the beginning of the 20^th^ century, the difference between their life expectancies has varied.

@@

These models were run separately for every combination of state, race, and sex. They were fit using two Markov chains of length 10,000. One thousand iterations remained after discarding the first two thousand iterations as burn-in and sampling every eighth iteration (to reduce autocorrelation). Each iteration produced death count estimates that were smoothed over time for every cause of death within each age group, which were used to estimate life expectancy, the life expectancy gap, and perform a cause of death decomposition. These 1,000 decompositions form the posterior distribution from which the credible intervals were estimated using the 2.5^th^ and 97.5^th^ percentiles.

@@

To describe the national trend in each causes' contribution to the life expectancy gap over time, we used random effects meta regression to model each state's mean contribution estimate from the previous step as a function of a state random effect and a linear spline for year (estimated using four internal knots). The model was inversely-weighted by the standard error of the mean estimate in each state-year. To identify divergent states, we compared the state-specific trends estimate in the previous step to the national pattern estimate by setting the random effect to zero.

@@

Cardiovascular disease (CVD) is the most common cause of death for both men and women of both races. Mortality from CVD decreased markedly since 1969 for both sexes and races (**eFigures 1 and 2**). However, white males underwent relatively larger mortality decreases than black males, resulting in a modest increase to the contribution of CVD to the life expectancy gap for most states (**Figure 3**). For females, black women experienced relatively larger mortality decreases over time compared to white females, resulting in a decrease to the CVD contribution to the life expectancy gap that was greater than one year in several states (**Figure 4**). For both genders, Michigan, Wisconsin, California, and DC exhibit recent contributions that are relatively larger than the general pattern, while Massachusetts and New York exhibit persistently lower levels of CVD contributions.

@@

For both races and both sexes, mortality from cancer peaked in the mid 1990s and decreased thereafter (**eFigures 3 and 4**). Among males, the peak mortality rate was much higher among blacks, which resulted in an inverted-U shape pattern in the difference in mortality between black and white males and in cancer's contribution to the life expectancy gap (**Figure 5**). For females, while cancer-related mortality continues to be higher among blacks in many states, the inverted-U pattern in mortality is similar between blacks and whites resulting in a flat general pattern in the contribution of cancer to the life expectancy over time (**Figure 6**). For both genders, DC and Wisconsin exhibit the largest cancer contributions (and also exhibit high initial values among males, but not females), while New York and Massachusetts underwent large decreases in the amount that cancer contributed to the life expectancy gap.

@@

Mortality from non-communicable disease has been increasing over time for black and white males in most states, with DC as a notable exception (**eFigure X**). Recently, there has been a slight decrease in mortality rates, especially among black males. These trends are similar among females, except that black females exhibited a sharper decline in mortality in recent years, eliminating the mortality difference in several states (**eFigure Y**). As such, the contribution of non-communicable disease to the mortality gap was generally flat among men with a recent decrease, and exhibited a gradual decrease over time among women (**Figures 7 and 8**). As with previous causes, New York and Massachusetts exhibit decreasing contributions, and Nevada has also made substantial progress. Furthermore, while DC's mortality rates decreased, they still have the highest contribution of non-communicable disease to the gap.

@@

Communicable disease is the least most common cause of death, although the 1990s exhibited an enormous increase in mortality related to communicable disease at the height of the HIV/AIDS epidemic before antiretroviral therapy was available. This spike was greater among males, especially blacks, and within states with larger urban centers. Thus, the contribution of communicable disease to the life expectancy gap exhibits a peak in the mid 1990s, with states diverging from this general pattern according to the intensity of their local epidemics (**Figures 11 and 12**). DC still exhibits higher communicable disease-related contributions among men, while New Jersey and Florida additionally show higher contributions among females.

@@

Mortality from all other causes generally decreased earlier in the time period, especially among blacks. White men and women have shown increasing mortality rates from all other causes since the 1980s, along with black women, while black men have continued to experience decreasing rates within several states. As such, the contribution of all other causes to the life expectancy gap has generally decreased over time, and reached equality in New York, Massachusetts, and Rhode Island for both genders (**Figures 13 and 14**). The most disparate states include Pennsylvania and DC, with much larger contributions estimated for all other causes in the earlier years within Pennsylvania and in the later years within DC.

@@

###Discussion 
- the next is to investigate potential state level policy and impact on underlying risk factors
- our analysis highlights a few specific causes in a few specific period
-injuries his CVD, not so much on cancer - because there is more homo. across contribution of cancer
- limitations - major groupings but they have lots of heterogeneity of causes within the grouping
- cont of HIV increases very dramatically, but the pace of decline differed - the places where there is a lag in the decline was the more rural places

@@

- While we found that many state-level trends followed the same pattern of COD-specific contributions, several states diverged in ways that either resulted in the near elimination of the life expectancy gap, or a less favorable result. New York and Massachusetts have virtually eliminated the black-white life expectancy gap for both men and women. While these states showed contribution decreases for all major causes of death except communicable disease, both states began with only a minimal contribution of CVD to the mortality gap.

@@

We used data over 45 years, across two race/ethnicities, and for all age groups. By examining a wide time window, trends were examined that provided context for the contributions estimated in the most recent year. This allowed us to recognize when the estimated contributions for a particular cause were relatively low *because they started low* vs. low *because they had decreased over time*. Furthermore, our analysis was centered on decomposing life expectancy at birth, a measure that integrates mortality rates across all ages. This is especially important in light of recent studies that have received criticism for narrowly focusing on a specific age group, which may present a distorted reality [Case & Deaton, Minton IJE 2016, ]. Alongside the decomposition results and age-standardized mortality rates, the scope of our analysis allows readers to examine differences in state-level inequality from many angles. [Methodologically, using the Bayesian autoregressive model, allowed us to estimate mortality rates in strata that were sparsely populated...*something about methods?*]

Our analysis has two main limitations. Firstly, we did not account for Hispanic ethnicity, implying that both black and white populations in our study include some portion of Hispanics that is known to differ over time and by state. We decided against excluding or adjusting for Hispanic ethnicity in this study because the indicator for ethnicity has been noted to have varying accuracy across states and during the earlier time period [ref]. Thus, states with demographic shifts due to immigration from Mexico such as Florida, California, and Texas, will have estimates of life expectancy that are likely higher than they would have been had Hispanics been excluded [ref]. In a previous sensitivity analysis, we found that accounting for Hispanic ethnicity did not have a substantial effect on the results [ref Harper 2012], providing some evidence that not accounting for Hispanic ethnicity should not substantially affect our findings.

Secondly, our analysis does not account for immigration, or migration between states. Individuals who immigrate to the US have been shown to have life expectancies at birth that are 5.8 and 4.8 years longer for men and women, respectively in 2008-2010 [ref Preston 2014]. As we are examine the difference in life expectancy between blacks and whites, this gap will be affected by the amount of immigration to the extent that new immigrants identify as black or white. Migration between states can also affect life expectancy. Migration is particularly important for the District of Columbia, because of it's small geographical bounds (it is easier to be considered migration when regions are smaller - how to say this?), and substantial in-migration from New York, in- and out-migration to contiguous Maryland and Virginia. Middle and upper-class black residents have been shown to move to the suburbs in Maryland and Virginia, while affluent whites are migrating to DC. These changes would both serve to increase the racial gap in DC.

helpful: https://www.nytimes.com/2014/08/16/upshot/mapping-migration-in-the-united-states-since-1900.html
-movement of southern blacks to northern states, and migration out of rust belt.
-DC - influx in immigrants since 1980s, huge migration from other US states (NY, MD then Virginia), but also huge out flux to MD and Virginia. Should almost consider these three places as one.
-lots of migration into MD
-NY - 24% are immigrants
-Texas has influx of immigrants to 17%
-lots of migration into Virginia, as well as immigration
-Florida is the state most known for in-migration, both from states (Northeast particularly NY) and from outside the US
-Alabama gets migrants from other southern states
-Connecticut - influx in those immigrating from outside US since 1990s

@@

Read: racial disparity in where people from DC move:
-https://ggwash.org/view/38326/theres-a-big-racial-disparity-in-where-people-born-in-dc-end-up-moving-or-not-moving
